### 2021-11-25 draft.tex
Deploying to gh-pages from @ gottacatchenall/thesis_proposal@4d7e4022ce3e32b44c8b4320cbf5baf230976336 🚀

Within the last several hundred years, human activity has induced rapid
changes in Earth's atmosphere, oceans, and surface. Greenhouse gas
emissions have caused an increase the temperature of both Earth's
terrain and oceans, and both agricultural and urban development has
rapidly reshaped Earth's land cover. The bulk of this change has
occurred within the last several hundred years, a geological instant,
inducing a sudden shift in conditions to Earth's climate and biosphere.
As a result \emph{ecological forecasting}---modeling how ecosystems and
their services will change in the future---and then using these
forecasts to make decisions to mitigate the negative consequences of
this change on ecosystems, their functioning, and the services they
provide to humans has emerged as an imperative for ecology and
environmental science (Dietze 2017). However, robust prediction of
ecological processes is, to say the least, quite difficult (Beckage
\emph{et al.} 2011; Petchey \emph{et al.} 2015). This difficultly is
compounded by a few factors, the first being that sampling ecosystems is
not easy. Ecological data is often biased, noisey, and sparse in both
space and time. The current paucity of ecological data has resulted in
much interest in developing global systems for \emph{ecosystem
monitoring} (Makiola \emph{et al.} 2020), which would systematize the
collection of biodiversity data in a manner that makes detecting and
predicting change more possible than at the moment (Urban \emph{et al.}
2021).

The second major challenge in ecological forecasting is that the
underlying dynamics of ecological processes are unknown and instead must
be inferred from this (sparse) data. Much of the history of
quantitatively modeling ecosystems have been done in the language of
dynamical systems, describing how the value of an observable state of
the system, represented by a vector of numbers
\([x_1, x_2, \dots, x_n]^T = \vec{x}\) changes over time, yielding
models in the form of differential equations in continuous-time
settings, \(\frac{dx}{dt} = f(x)\), or difference equations in
discrete-time settings, \(x_t = f(x_{t-1})\), where
\(f:\mathbb{R}^n \to \mathbb{R}^n\) is an arbitrary function describing
how the system changes on a moment-to-moment basis (e.g.~in the context
of communities, \(f\) could be Lotka-Voltera, Holling-Type-II or
DeAngelis-Beddington functional response). The form of this functional
response in real systems, and whether it is meaningfully non-zero for a
given species interaction, is effectively unknown and must be predicted
(Strydom \emph{et al.} 2021b), and some forms of these dynamics are
inherently more ``forecastable'' than others (Beckage \emph{et al.}
2011; Chen \emph{et al.} 2019; Pennekamp \emph{et al.} 2019). The
initial success of these forms of models can be traced back to the
larger program of ontological reductionism, which became the default
approach to modeling in the sciences after its early success in physics,
which, by the time ecology was becoming a quantitative science (sometime
in the 20th century, depending on who you ask), became the foundation
for mathematical models in ecology.

@@

But these problems are not solely unique to ecology. The term
\emph{ecological forecasting} implicitly creates an analogy with weather
forecasting. Although it has become a trite joke to complain about the
weather forecast being wrong, over the least 50 years the field of
numerical weather prediction (NWP) has dramatically improved our ability
to predict weather across the board (Bauer \emph{et al.} 2015). The
success of NWP, and the Earth observation systems that support it (Hill
\emph{et al.} 2004), should serve as a template for development of a
system for monitoring Earth's biodiversity. Much like ecology, NWP is
faced with high-dimensional systems that are governed by different
mechanisms at different scales. The success of NWP is that, rather than,
say, attempt to forecast the weather in Québec by applying Navier-Stokes
to entire province, to instead use simulation models which describe
known mechanisms at different scales, and use the availability to
increasing computational power to directly simulate many batches of
dynamics which directly incorporate stochasticity and uncertainty in
parameter estimates via random number generation.

But forecasting is only half the story---if indeed ``{[}ecologists{]}
have hitherto only interpreted the world in various ways; the point is
to change it,'' then once we have a forecast about how an ecosystem will
change in the future, what if this forecast predicts a critical
ecosystem service will deteriorate? We are still left with the question,
what do we in the time being to mitigate the potentially negative
consequences a forecast predicts? In this framing, mitigating the
consequences of anthropogenic change on ecosystems becomes an
optimization problem: given a forecast of the future state of the
system, and some ``goal'' state for the future, the problem is then to
optimize our intervention into the system to maximize the probability
the system approaches our ``goal'' state. This dissertation aims to
provide a framework for ecosystem monitoring and forecasting
(fig.~\ref{fig:thesis}, left), and each chapter address some aspect of
this pipeline to data from a monitoring network to forecasts to
mitigation strategy (fig.~\ref{fig:thesis}, right).

@@

The primary research challenges this thesis addresses are: how do we
design ecological samples to? How do we build the software
infrastructure to assimilate data from a variety of sources? And how do
we propagate uncertainty from data to forecasts? The flow of chapters
follows the flow in fig.~\ref{fig:thesis} (left), from data collection
via a monitoring network, to forecasting an essential biodiversity
variable (EBV), to optimizing mitigation strategy based on constraints.
In chapter one, we discuss how simulation can aid in the design of
ecological samples and monitoring network design. In chapter two we use
data to forecast the uncoupling of a plant-pollinator network. In
chapter three, we apply simulation methods in landscape ecology to
optimize corridor placement to maximize the time-until-extinction of a
metapopulation. The fourth and final chapter is the software
(\emph{MetacommunityDynamics.jl}) which enables the rest of the
dissertation.

@@

This chapter uses simulation models to investigate the relationship
between species relative abundance, sampling effort, and probability of
observing an interaction between species in order to aid in the design
of samples of ecological interactions across space, and to provide a
null expectation of the false-negative probability for a dataset of a
given size. Further it then proposes a method for optimizing the spatial
sampling locations to maximize the probability of detecting an
interaction between two species given a fixed number of total of
observations and the distributions of each species. This addresses the
optimization of monitoring network part of the flow from data to
mitigation at the top of fig.~\ref{fig:thesis}, left. There is more than
one way to observe a false-negative when sampling interactions---
fig.~\ref{fig:fnrtaxonomy} shows a taxonomy of false-negatives in
occurrence, co-occurrence, and interaction data. The goal of this
chapter is to untangle the relationship between co-occurrence
false-negatives (and their relation to relative abundance) and
interaction false-negatives.

@@

\begin{figure}
\centering
\includegraphics{./figures/negativebinomial.png}
\caption{Relationship between total number of observed interactions
negatives (x-axis) and the probability that an interaction is a
true-negative as a function of different realized false-negative rates
(colors), assuming each observed negative is independent---the same as
the negative-bernoulli distribution.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{./figures/ch1.png}
\caption{Chapter One conceptual figure. Left: the sources of data and
how they can be combined. Right: The flow from data to interaction
prediction using a few different interaction prediction models.}
\end{figure}

@@

The biggest challenge in implementing simulated annealing in this
context is defining a proposal function for landscape modifications. At
the moment this is done by computing the minimum-spanning-tree (MST) of
the spatial nodes (locations where occurrence has been observed), and
then proposing corridors that only connect nodes that are already
connected in the MST. The primary reason for doing this is the cut down
the size of the solution space to enable quicker convergence, although
the final software that implements this algorithm will enable
alternative methods of proposing modifications.

>>>>

Within the last several hundred years, human activity has induced rapid
changes in Earth's atmosphere, oceans, and surface. Greenhouse gas
emissions have caused an increase in the temperature of both Earth's
terrain and oceans, and both agricultural and urban development has
rapidly reshaped Earth's land cover. The bulk of this change has
occurred within the last several hundred years, a geological instant,
inducing a sudden shift in conditions to Earth's climate and biosphere.
As a result \emph{ecological forecasting}---modeling how ecosystems and
their services will change in the future---and then using these
forecasts to make decisions to mitigate the negative consequences of
this change on ecosystems, their functioning, and the services they
provide to humans has emerged as an imperative for ecology and
environmental science (Dietze 2017). However, robust prediction of
ecological processes is, to say the least, quite difficult (Beckage
\emph{et al.} 2011; Petchey \emph{et al.} 2015). This difficultly is
compounded by a few factors, the first being that sampling ecosystems is
not easy. Ecological data is often biased, noisey, and sparse in both
space and time. The current paucity of ecological data has resulted in
much interest in developing global systems for \emph{ecosystem
monitoring} (Makiola \emph{et al.} 2020), which would systematize the
collection of biodiversity data in a manner that makes detecting and
predicting change more possible than at the moment (Urban \emph{et al.}
2021).

The second major challenge in ecological forecasting is that the
underlying dynamics of ecological processes are unknown and instead must
be inferred from this (sparse) data. Much of the history of
quantitatively modeling ecosystems have been done in the language of
dynamical systems, describing how the value of an observable state of
the system, represented by a vector of numbers
\([x_1, x_2, \dots, x_n]^T = \vec{x}\) changes over time, yielding
models in the form of differential equations in continuous-time
settings, \(\frac{dx}{dt} = f(x)\), or difference equations in
discrete-time settings, \(x_t = f(x_{t-1})\), where
\(f:\mathbb{R}^n \to \mathbb{R}^n\) is an arbitrary function describing
how the system changes on a moment-to-moment basis (e.g.~in the context
of communities, \(f\) could be Lotka-Volterra, Holling-Type-II or
DeAngelis-Beddington functional response). The form of this functional
response in real systems, and whether it is meaningfully non-zero for a
given species interaction, is effectively unknown and must be predicted
(Strydom \emph{et al.} 2021b), and some forms of these dynamics are
inherently more ``forecastable'' than others (Beckage \emph{et al.}
2011; Chen \emph{et al.} 2019; Pennekamp \emph{et al.} 2019). The
initial success of these forms of models can be traced back to the
larger program of ontological reductionism, which became the default
approach to modeling in the sciences after its early success in physics,
which, by the time ecology was becoming a quantitative science (sometime
in the 20th century, depending on who you ask), became the foundation
for mathematical models in ecology.

@@

But these problems are not solely unique to ecology. The term
\emph{ecological forecasting} implicitly creates an analogy with weather
forecasting. Although it has become a trite joke to complain about the
weather forecast being wrong, over the last 50 years the field of
numerical weather prediction (NWP) has dramatically improved our ability
to predict weather across the board (Bauer \emph{et al.} 2015). The
success of NWP, and the Earth observation systems that support it (Hill
\emph{et al.} 2004), should serve as a template for development of a
system for monitoring Earth's biodiversity. Much like ecology, NWP is
faced with high-dimensional systems that are governed by different
mechanisms at different scales. The success of NWP is that, rather than,
say, attempt to forecast the weather in Québec by applying Navier-Stokes
to entire province, to instead use simulation models which describe
known mechanisms at different scales, and use the availability to
increasing computational power to directly simulate many batches of
dynamics which directly incorporate stochasticity and uncertainty in
parameter estimates via random number generation.

But forecasting is only half the story---if indeed ``{[}ecologists{]}
have hitherto only interpreted the world in various ways; the point is
to change it,'' then once we have a forecast about how an ecosystem will
change in the future, what if this forecast predicts a critical
ecosystem service will deteriorate? We are still left with the question,
what do we do in the time being to mitigate the potentially negative
consequences a forecast predicts? In this framing, mitigating the
consequences of anthropogenic change on ecosystems becomes an
optimization problem: given a forecast of the future state of the
system, and some ``goal'' state for the future, the problem is then to
optimize our intervention into the system to maximize the probability
the system approaches our ``goal'' state. This dissertation aims to
provide a framework for ecosystem monitoring and forecasting
(fig.~\ref{fig:thesis}, left), and each chapter addresses some aspect of
this pipeline to data from a monitoring network to forecasts to
mitigation strategy (fig.~\ref{fig:thesis}, right).

@@

The primary research challenges this thesis addresses are: how do we
design ecological samples to detect change? How do we build the software
infrastructure to assimilate data from a variety of sources? And how do
we propagate uncertainty from data to forecasts? The flow of chapters
follows the flow in fig.~\ref{fig:thesis} (left), from data collection
via a monitoring network, to forecasting an essential biodiversity
variable (EBV), to optimizing mitigation strategy based on constraints.
In chapter one, we discuss how simulation can aid in the design of
ecological samples and monitoring network design. In chapter two we use
data to forecast the uncoupling of a plant-pollinator network. In
chapter three, we apply simulation methods in landscape ecology to
optimize corridor placement to maximize the time-until-extinction of a
metapopulation. The fourth and final chapter is the software
(\emph{MetacommunityDynamics.jl}) which enables the rest of the
dissertation.

@@

This chapter uses simulation models to investigate the relationship
between species relative abundance, sampling effort, and probability of
observing an interaction between species in order to aid in the design
of samples of ecological interactions across space, and to provide a
null expectation of the false-negative probability for a dataset of a
given size. Further, it then proposes a method for optimizing the
spatial sampling locations to maximize the probability of detecting an
interaction between two species given a fixed number of total of
observations and the distributions of each species. This addresses the
optimization of monitoring network part of the flow from data to
mitigation at the top of fig.~\ref{fig:thesis}, left. There is more than
one way to observe a false-negative when sampling interactions---
fig.~\ref{fig:fnrtaxonomy} shows a taxonomy of false-negatives in
occurrence, co-occurrence, and interaction data. The goal of this
chapter is to untangle the relationship between co-occurrence
false-negatives (and their relation to relative abundance) and
interaction false-negatives.

@@

\begin{figure}
\centering
\includegraphics{./figures/negativebinomial.png}
\caption{Relationship between total number of observed interactions
negatives (x-axis) and the probability that an interaction is a
true-negative as a function of different realized false-negative rates
(colors), assuming each observed negative is independent---the same as
the negative-Bernoulli distribution.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{./figures/ch1.png}
\caption{Chapter two conceptual figure. Left: the sources of data and
how they can be combined. Right: The flow from data to interaction
prediction using a few different interaction prediction models.}
\end{figure}

@@

The biggest challenge in implementing simulated annealing in this
context is defining a proposal function for landscape modifications. At
the moment this is done by computing the minimum-spanning-tree (MST) of
the spatial nodes (locations where occurrence has been observed), and
then proposing corridors that only connect nodes that are already
connected in the MST. The primary reason for doing this is to the cut
down the size of the search space to enable quicker convergence,
although the final software that implements this algorithm will enable
alternative methods of proposing modifications.
